The Supervisionary proof-checking kernel

### 2021-10-08 paper/prisc-2022.tex
prisc-2022: some minor word rearranging, and squashing

The kernel also offers an abstract machine interface to user-space software, smoothing over the vulgarities of programming against any particular peripheral, say.
Useful abstractions, such as sockets and a file-system, are also commonly offered.
This interface is largely exposed as \deffont{system calls}, defining an \deffont{application binary interface}, or ABI henceforth, which user-space and kernel must understand to interact.
%Executing a system call forces a \deffont{context switch}, within which the thread of execution enters the kernel.
The result of a system call may be written directly into the memory of the calling code, via a pointer argument, for example.
Note the asymmetry: in theory, the kernel can read and modify all of a user-space program's state.

@@

The manipulation and querying of kernel objects is performed by the kernel itself on behalf of untrusted code.
Space constraints prevent us from describing the entire Supervisionary ABI for working with, and on, kernel objects.
However, note that theorems are also constructed in a similar way to terms, inductively building derivation trees.
For example, the HOL symmetry rule is exposed as:
\begin{displaymath}
\mathtt{Theorem.Handle.AllocateSym(pre, out)}
\end{displaymath}
Here, $\mathtt{pre}$ points-to a theorem $\Gamma \vdash r = s$ and after succeeding, passing obvious checks, $\mathtt{out}$ contains a handle that points-to a newly-allocated theorem $\Gamma \vdash s = r$.

@@

\deffont{Jailing}, wherein a process voluntarily sheds capabilities, is common.
This can be captured by allowing a change in the prevailing policy, $\phi$, to a new policy $\psi$, as long as one can prove that $\psi$ is a refinment of $\phi$: $\{\} \vdash \fall{k}\fall{u}\fall{s} \phi\ k\ u\ s \longrightarrow \psi\ k\ u\ s$.
Note that this expresses that the states described by $\psi$ are a subset of those described by $\phi$.

\deffont{Proof carrying code} modified the operating system loader, and endowed binaries with skeleton proofs of adherence to some policy, later reconstructed by the loader prior to execution.
The ideas sketched above are different: proofs are generated dynamically, as the program executes, and perhaps more aptly called \deffont{proof generating code}.

@@

Switching tack, proof-assistants in the LCF lineage are also architected around a \deffont{kernel}, with ostensibly little in common with the kernels discussed so far.
A closer examination shows that both are tasked with preserving important invariants in the face of unbounded interaction with untrusted code.
Moreover, whilst operating system kernels enjoy a monopoly on direct hardware access, proof-assistant kernels enjoy a monopoly on the \deffont{authentication} of theorems.
In short, both act as the roots of all trust within their domains.
Yet, the two are implemented in different ways, with LCF kernels implemented as libraries in a \deffont{meta-language}, usually a functional language.
Kernel interaction is via an \emph{API}, rather than an ABI.

@@

Supervisionary is an extended exercise in pointer manipulation.
The semantics of our system calls can therefore be expressed as Hoare Triples, using formulae of \deffont{Separation Logic} as an assertion language.
Write $\mathtt{h} \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)}$ to assert that the handle $\mathtt{h}$ points-to a term application (of the term pointed-to by $l$ to the term pointed-to by $r$).
Write $\mathtt{out} \mapsto \mathtt{b}$ to assert that $\mathtt{out}$ points-to the Boolean value $\mathtt{b}$.
Write $P \star Q$ for the \deffont{separating conjunction}.
Finally, write $\{ P \} C \{ Q \}$ to assert that $C$ executing in a state satisfying $P$ succeeds and produces a state satisfying $Q$.
We then have, for example:
\begin{gather*}
\{ h \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)} \star \mathtt{out} \mapsto b \} \\
\mathtt{Term.Handle.IsApplication(h, out)} \\
\{ h \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)} \star \mathtt{out} \mapsto \mathtt{True} \}
\end{gather*}
Note that the structural \deffont{frame rule} captures the fact that any system call only manipulates a defined footprint of kernel and untrusted code's memory.

>>>>

The kernel also offers an abstract machine interface to user-space software, smoothing over the vulgarities of programming against any particular peripheral, say.
Useful abstractions, such as sockets and a filesystem, are also commonly offered.
This interface is largely exposed as \deffont{system calls}, defining an \deffont{application binary interface}, or ABI henceforth, which user-space and kernel must understand to interact.
%Executing a system call forces a \deffont{context switch}, within which the thread of execution enters the kernel.
The result of a system call may be written directly into the memory of the calling code, via a pointer argument, for example.
Note the asymmetry: in theory, the kernel can read and modify all of a user-space program's state.

@@

The manipulation and querying of kernel objects is performed by the kernel itself on behalf of untrusted code.
Space constraints prevent us from describing the entire Supervisionary ABI for working with, and on, kernel objects.
However, note that theorems are also constructed in a similar way to terms, inductively building derivation trees.
For example, the HOL symmetry rule is exposed as:
\begin{displaymath}
\mathtt{Theorem.Handle.AllocateSym(pre, out)}
\end{displaymath}
Here, $\mathtt{pre}$ points-to an existing theorem object $\Gamma \vdash r = s$ and after succeeding, passing obvious checks, $\mathtt{out}$ contains a handle that points-to a newly-allocated theorem $\Gamma \vdash s = r$.

@@

\deffont{Jailing}, wherein a process voluntarily sheds capabilities, is common.
This can be captured by allowing a change in the prevailing policy, $\phi$, to a new policy $\psi$, as long as one can prove that $\psi$ is a refinment of $\phi$: $\{\} \vdash \fall{k}\fall{u}\fall{s} \psi\ k\ u\ s \longrightarrow \phi\ k\ u\ s$.
Note that this expresses that the states described by $\psi$ are a subset of those described by $\phi$.

\deffont{Proof carrying code} modified the operating system loader, and endowed binaries with skeleton proofs of adherence to some policy, later reconstructed by the loader prior to execution.
The ideas sketched above are different: proofs are generated dynamically, as the program executes, and more aptly called \deffont{proof generating code}.
Moreover, theorems could be generated statically, offline, and used dynamically:

@@

Proof-assistants in the LCF lineage are also architected around a \deffont{kernel}, with ostensibly little in common with operating system kernels.
Examining, one sees that both are tasked with preserving invariants in the face of unbounded interaction with untrusted code.
Moreover, whilst operating system kernels enjoy a monopoly on hardware access, proof-assistant kernels enjoy a monopoly on theorem \deffont{authentication}.
In short, both act as roots of all trust within their domains.
Yet, the two are implemented in different ways, with LCF kernels implemented as libraries in a \deffont{meta-language}, usually a functional language.
Kernel interaction is via an \emph{API}, rather than an ABI.

@@

Supervisionary is an extended exercise in pointer manipulation.
The semantics of our system calls can therefore be expressed as Hoare Triples, using formulae of \deffont{Separation Logic} as an assertion language.
Write $\mathtt{h} \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)}$ to assert that the handle $\mathtt{h}$ points-to a term application (of the term pointed-to by $l$ to the term pointed-to by $r$).
Write $\mathtt{out} \mapsto \mathtt{b}$ to assert that $\mathtt{out}$ points-to the Boolean value $\mathtt{b}$.
Write $P \star Q$ for the \deffont{separating conjunction}.
Finally, write $\{ P \} C \{ Q \}$ to assert that $C$ executing in a state satisfying $P$ succeeds and produces a state satisfying $Q$.
We then have the following example \deffont{local axiom}:
\begin{gather*}
\{ h \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)} \star \mathtt{out} \mapsto b \} \\
\mathtt{Term.Handle.IsApplication(h, out)} \\
\{ h \mapsto_{\mathtt{trm}} \mathtt{Application(l, r)} \star \mathtt{out} \mapsto \mathtt{True} \}
\end{gather*}
Note that the structural \deffont{frame rule} captures the fact that any system call only manipulates a defined footprint of kernel and untrusted code's memory.
